Robotic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-09-16 main.tex
Started work on the discussion section.

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage[natbib=true,style=authoryear]{biblatex}
\addbibresource{references.bib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage{todonotes}

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. In the speed range tested, the commanded
steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
proportional feedback gain linearly with respect to speed stabilizes the
normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
Figure~\ref{fig:eigenvalues}.
\todo[inline]{Could add the experimental weave mode identification to this
figure and consider lowering the gain on the model to make the model match the
experimental values.}
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the bicycle without the inertial effect of a rigid rider (left) and
    with a rigid rider (right). Vertical blue lines indicate the two speeds we
    perform experiments at. Grey shaded region is the linear stable speed
    range. Geometric and inertial parameters for these plots were estimated
    with the methods presented in \cite{Moore2012} and software packages
    BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015}.}
  \label{fig:eigenvalues}
\end{figure}

@@

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a pertubation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance assist state on and off while keeping all other explanatory
factors at their centered mean values. This figure is created by setting all
explanatory variables to there mean and calculating the probality from
Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance-assist
system halves (0.53) the odds that a perturbation results in a fall. This
figure shows that for relatively large impulses the probability to fall is
unity for both states of balance assist on and off. And for relatively small
impulses the probability to fall is null for both states. But for impulses in
the magnitude region (-1 to 0.5 STD), i.e. around the mean-centered fall
threshold, the probability of falling is significantly lowered with the balance
assist system on. The skewedness of the probaility curves arrives from the
interaction effects.\todo{check this statement about skewedness this and think
about it} Figure~\ref{fig:probablity-10kph} shows the same result for the
10~\kph trials which has a similar trend of reducing the probability to fall
with the balance assist system turned on, but the effect is not significant.
%
\begin{figure}
  \centering
  % NOTE : We could uncomment the longer subcaption if titles removed from
  % figures. It is currently redundant.
  \subcaptionbox{%Fall probability for the 6~\si{\kph} trials.
    \label{fig:probability-6kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{%Fall probability for the 10~\si{\kph} trials.
    \label{fig:probability-10kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance-assist on
    (blue) or off (red) when all other predictor variables are fixed at zero,
    except for the interaction effect between the balance-assist and the
    angular impulse. The ordinate axis is standard deviation around the mean
    (all perturbations), centered per participant.}
  \label{fig:probability}
\end{figure}

\section{Discussion}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain
some insight.
But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall
(MCiO=0.53)~\todo{Make a variable for MCiO}.
This means if the odds are a 1000:1, turning on the balance-assist system
reduces the odds to 500:1.
However, the probability that a fall occurs is only reduced from $0.999$ to
$0.998$ in that case.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance-assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse.
This means that the balance assist is most effective for perturbations close to
the subject's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

\todo{Consider making a specific value of calcualting the probabilty for a set
of inputs to the regression model.}

To illustrate the effect of the balance-assist system on fall probability, we will
give an example of how the data collected during the experiments is used to predict
fall probability. We use \ref{eq:log-regress} and the coefficients in 
\ref{tab:freq-coefs-6}. For simplicities sake, the interaction effects are not included.
Let's assume that the mean angular impulse $\bar{L}$ of all the perturbations applied to 
a participant is 100~\si{\newton}, and the standard deviation $\sigma^{L}=15$. The centred
and scaled angular impulse can be calculated by substracting $\bar{L}$ from the applied
angular impulse $L$, and dividing this by $\sigma^{L}$. The same applies for the 
perturbation order $c$, initial roll angle $\phi_0$ and initial steer angle $\delta_0$.
If we take the coefficients estimated for cycling at 6~\si{\kilo\meter\per\hour}, the
log-odds of falling can be calculated as follows.

\begin{align}
	\log \left( \frac{p_{ij}}{1-p_{ij}} \right) & = \beta + \sum_{k}^{k=0}\alpha_k 
	\frac{x_{ij}^{k}-\bar{x_{ij}^{k}}} {\sigma^{x^{k}}} 
	& = -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}} 
	-0.77\cdot\frac{10-20}{11.54} 
	-0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}} 
	-0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}}
	-0.64\cdot s
	& = 1.42 -0.64 \cdot s
\end{align}

@@

Angular impulse magnitude has the largest significant effect for predicting
fall probability, Tables~\ref{tab:freq-coefs-6} and \ref{tab:freq-coefs-10}.
An increase in angular impulse increases the fall probability both at 6 and
10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds is
approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a more
important predictor at higher speeds compared to lower speeds. The reason for
this likely has to do with the width of the treadmill and may also be why
balance assist system did not have a statistically significant effect at
10~\si{\kph}. As a bicycle travels at higher speeds, the same perturbation
causes larger lateral deviations. At 10~\si{\kph} almost all falls were due to
the bicycle exiting the maximum width of the treadmill. If the same experiment
was performed on an infinite plane, the riders may have recovered from more
perturbations. At 6~\si{\kph} the riders could often recover in the allotted
treadmill width. Our results are very much dependent on the two modes of
falling with use: exit the treadmill width or foot is placed on the belt. Cycle
paths are a similar width as the treadmill, so rider's are often limited in
width when recovering from a fall.

@@

An increase in the number of perturbation that a participant already
experienced, decreases the probability that a fall occurs. This is likely due
to the participants learning how to better recover from the perturbation over
the duration of the experiment. This effect is significant at both 6 and
10~\si{\kph}, and in the same order of magnitude. This suggests that the
learning effect that occurs during the experiment is not strongly dependent on
the speed.

@@

Difficult to layer this onto the fall statistics, because we don't have the
details of how people fall. If such data existed we could make estimates on the
number of falls reduced if everyone road such a bike.

@@

% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\printbibliography

>>>>

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage[natbib=true,style=authoryear]{biblatex}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\addbibresource{references.bib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage{todonotes}

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. In the speed range tested, the commanded
steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
proportional feedback gain linearly with respect to speed stabilizes the
normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
Figure~\ref{fig:eigenvalues}.
\todo[inline]{Could add the experimental weave mode identification to this
figure and consider lowering the gain on the model to make the model match the
experimental values.}
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the bicycle without the inertial effect of a rigid rider (left) and
    with a rigid rider (right). Vertical blue lines indicate the two speeds we
    perform experiments at. Grey shaded region is the linear stable speed
    range. Geometric and inertial parameters for these plots were estimated
    with the methods presented in \cite{Moore2012} and software packages
    BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015}.}
  \label{fig:eigenvalues}
\end{figure}

@@

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a pertubation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance assist state on and off while keeping all other explanatory
factors at their centered mean values. This figure is created by setting all
explanatory variables to there mean and calculating the probality from
Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance-assist
system halves (0.53) the odds that a perturbation results in a fall. This
figure shows that for relatively large impulses the probability to fall is
unity for both states of balance assist on and off. And for relatively small
impulses the probability to fall is null for both states. But for impulses in
the magnitude region (-1 to 0.5 STD), i.e. around the mean-centered fall
threshold, the probability of falling is significantly lowered with the balance
assist system on. The skewedness of the probaility curves arrives from the
interaction effects.\todo{check this statement about skewedness this and think
about it} Figure~\ref{fig:probability-10kph} shows the same result for the
10~\si{\kph} trials which has a similar trend of reducing the probability to
fall with the balance assist system turned on, but the effect is not
significant.
%
\begin{figure}
  \centering
  % NOTE : We could uncomment the longer subcaption if titles removed from
  % figures. It is currently redundant.
  \subcaptionbox{%Fall probability for the 6~\si{\kph} trials.
    \label{fig:probability-6kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{%Fall probability for the 10~\si{\kph} trials.
    \label{fig:probability-10kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance-assist on
    (blue) or off (red) when all other predictor variables are fixed at zero,
    except for the interaction effect between the balance-assist and the
    angular impulse. The ordinate axis is standard deviation around the mean
    (all perturbations), centered per participant.}
  \label{fig:probability}
\end{figure}

\section{Discussion}
%
We have shown that at 6~\si{\kph} the addition of balance assist control
reduces the chance that a rider will fall when perturbed around the limits of their
control authority. But this effect diminishes at the higher speed
scenario of 10~\si{\kph}. We were only able to test these two speed-gain
scenarios for mostly homogeneous sets of riders within the resources of this
research project, but additional experimental work could help understand more
completely the range and limits of the positive effect of the balance system.
For example, it is possible that simply increasing the controller gain at
10~\si{\kph} also results in a significant positive effect.

\subsection{Interpretation of the Results}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain
some insight.
But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall:
\(e^{\alpha_k}=0.53\).
This means if the odds are a 1000:1, turning on the balance-assist system
reduces the odds to 500:1.
However, in that case the probability that a fall occurs is only reduced from
$0.999$ to $0.998$.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance-assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse.
This means that the balance assist is most effective for perturbations close to
the subject's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

To illustrate the effect of the balance-assist system on fall probability, we
will give an example of how the data collected during the experiments is used
to predict fall probability. We use \ref{eq:log-regress} and the coefficients
in \ref{tab:freq-coefs-6}. For simplicities sake, the interaction effects are
not included.  Let's assume that the mean angular impulse $\bar{L}$ of all the
perturbations applied to a participant is 100~\si{\newton}, and the standard
deviation $\sigma^{L}=15$. The centred and scaled angular impulse can be
calculated by subtracting $\bar{L}$ from the applied angular impulse $L$, and
dividing this by $\sigma^{L}$. The same applies for the perturbation order $j$,
initial roll angle $\phi_0$, and initial steer angle $\delta_0$.  If we take
the coefficients estimated for cycling at 6~\si{\kph}, the log-odds of falling
can be calculated as follows.
%
\begin{align}
  \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
  =
  \beta + \sum_{k}^{k=0}\alpha_k \frac{x_{ij}^{k}-\bar{x_{ij}^{k}}}{\sigma^{x^{k}}}
  =
  & -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}}
    -0.77\cdot\frac{10-20}{11.54}  \\
  & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot s \\
  =
  & 1.42 -0.64 \cdot s
\end{align}

@@

\subsection{Stability and Human Controlled Plant Dynamics}
%
The linear Carvallo-Whipple model indicates that the steer controller
stabilizes the bicycle-rider system, but this model assumes the rider's hands
are not connected to the handlebars and that they clamp their body as rigidly
as possible to the rear frame. In reality, the system's behavior is likely more
akin to a marginally stable or an easily controllable unstable system due to
the various un-modeled effects. Our system may not result in a definitely
stable system, i.e. cannot fall, but having plant eigenvalues with very small
unstable eigenvalue real parts correlates to ease of control~\citep{Hess2012}.

The controller design we utilize,
Equation~\ref{eq:implemented-controller}, also increases the weave mode
frequency by a factor of about three up to about 1~\si{\hertz}. This bandwidth
is still controllable by the human's neuromuscular system, but may feel
unnatural as it is more akin to what the steering would feel like at in the
\SIrange{30}{40}{\kph} speed range. \Citet{Hanakam2023} reported
dissatisfaction in subjective rider feeling on their similar bicycle to ours
and this effect to the human-controlled plant dynamics could be connected to
this.

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why balance assist system did not have a statistically
significant effect at 10~\si{\kph}. As a bicycle travels at higher speeds, the
same perturbation magnitude causes larger lateral deviations. At 10~\si{\kph}
almost all falls were due to the bicycle exiting the maximum width of the
treadmill. If the same experiment was performed on an infinitely wide plane,
the riders may have recovered from more perturbations. At 6~\si{\kph} the
riders could often recover in the allotted treadmill width due to the smaller
lateral deviations. Our results are very much dependent on the two modes of
falling with use: exit the treadmill width or foot is placed on the belt. Cycle
paths are a similar width as the treadmill, so rider's are often limited in
width when recovering from a fall.

@@

\subsection{Learning Effect}
%
An increase in the number of perturbation that a participant already
experienced, decreases the probability that a fall occurs. This is likely due
to the participants learning how to better recover from the perturbation over
the duration of the experiment. This effect is significant at both 6 and
10~\si{\kph}, and in the same order of magnitude. This suggests that the
learning effect that occurs during the experiment is not strongly dependent on
the speed.

\subsection{Non-significant Predictors}

@@

\subsection{Extrapolation to Natural Falls}
%
The positive effect of the balance assist system is coupled to the assumptions
and experimental scenarios we implemented and there is unfortunately no simple
way to extrapolate our results to reductions of single-actor crashes we may see
if such a system were deployed widely to bicyclists. Although, our results do
indicate that we would see such a reduction, even if only in a class of
single-actor crashes that most resemble our experimental design. If there were
more comprehensive and detailed natural data of how people fall we could make
estimates on the number of falls reduced if everyone road a balance assist
bicycle.

@@

\printbibliography
